---
title: "BSG-MDS Practical 2 Statistical Genetics"
author: "Biel Caballero and Gerard Gomez"
date: "07/11/2023, submission deadline 14/11/2023"
output:
  pdf_document: default
  html_document: default
---

```{r,echo=FALSE}
library(data.table)
library(genetics)
library(HardyWeinberg)
```

#Hardy Weinberg Equilibrium

##Create dataset
```{r}
data<-fread("TSIChr22v4.raw", header = TRUE)
genetic_data<-data[,-c(1:6)]
```

### 1. How many variants are there in this database? What percentage of the data is missing?

```{r}

```

### 2. Calculate the percentage of monomorphic variants. Exclude all monomorphics from the database for all posterior computations of the practical. How many variants do remain in your database?

```{r}

```

### 3. Extract polymorphism rs587756191_T from the data, and determine its genotype counts. Apply a chi-square test for Hardy-Weinberg equilibrium, with and without continuity correction. Also try an exact test, and a permutation test. You can use thee functions HWChisq, HWExact and HWPerm for this purpose. Do you think this variant is in equilibrium? Argue your answer.

```{r}

```

### 4. Determine the genotype counts for all polymorphic variants, and store them in a p × 3 matrix.

```{r}

```

### 5. Apply an exact test for Hardy-Weinberg equilibrium to each SNP. You can use function HWExactStats for fast computation. What is the percentage of significant SNPs (use α = 0.05)? Is this the number of markers that you would expect to be out of equilibrium by the effect of chance alone?

```{r}

```

### 6. Which SNP is most significant according to the exact test results? Give its genotype counts. In which sense is this genotypic composition unusual?

```{r}

```

### 7. Compute the inbreeding coefficient (f) for each SNP, and make a histogram of f. You can use function HWf for this purpose. Give descriptive statistics (mean, standard deviation, etc) of f calculated over the set of SNPs. What distribution do you expect f to follow theoretically? Use a probability plot to confirm your idea.

```{r}

```

### 8. Apply the exact test for HWE to each SNP, using different significant levels. Report the number and percentage of significant variants using an exac test for HWE with α = 0.10, 0.05, 0.01 and 0.001. State your conclusions.

```{r}

```







