---
title: "BSG-MDS Practical 3 Statistical Genetics"
author: "Biel Caballero and Gerard Gomez"
date: "05/12/2023, submission deadline 12/12/2023"
output:
  html_document: default
  pdf_document: default
---

# Population substructure

### 1. Perform the alleles test for this data set. Provide the p-value and the odds ratio and comment on the results.
```{r}
df <- matrix(c(112,278,150,206,348	,150),byrow=TRUE,ncol=3)
colnames(df) <- c("AA","Aa","aa")
rownames(df) <- c("Cases","Controls")
df

alleles<-cbind(2*df[,1]+df[,2],2*df[,3]+df[,2])
colnames(alleles) <- c("A","a")
fisher.test(alleles)
```
The p-value is 0.000231 and the odds ratio is 0.7406221. This means that we accept the alternative hypothesis meaning that the true odds ratio is not equal to 1.

### 2. Test for association using a codominant, a dominant and a recessive model. Provide the p-values for all the tests and comment on the results.
```{r}
#Codominant
codominant <- matrix(c(112,278,150,206,348,150),byrow=TRUE,ncol=3)
colnames(codominant) <- c("AA","AB","BB")
rownames(codominant) <- c("Cases","Controls")
codominant

results<-chisq.test(codominant)
results
results$expected
fisher.test(codominant)

#Dominant
dominant <- cbind(codominant[,1],codominant[,2]+codominant[,3])
colnames(dominant) <- c("AA","AB or BB")
rownames(dominant) <- c("Cases","Control")
dominant
results<-chisq.test(dominant)
results
results<-chisq.test(codominant, correct = FALSE)
results

#Recessive
recessive <- cbind(codominant[,1]+codominant[,2],codominant[,3])
colnames(recessive) <- c("AA or AB","BB")
rownames(recessive) <- c("Cases","Control")
recessive
results<-chisq.test(recessive)
results
fisher.test(recessive)
```

### 3. Plot the risk of disease as a function of the number of m alleles. Comment on the results. Which model seems most appropriate?
```{r}

```

### 4. Perform Armitage trend test for this data set. Does the null hypothesis β1 = 0 hold? Comment on your response.
```{r}

```

### 5. Is there evidence for association of this marker with the disease? Argument your response
```{r}

```







