---
title: "BSG-MDS Practical 3 Statistical Genetics"
author: "Biel Caballero and Gerard Gomez"
date: "05/12/2023, submission deadline 12/12/2023"
output:
  html_document: default
  pdf_document: default
---

```{r, echo=FALSE}
library(ggplot2)
```


# Population substructure

### 1. Perform the alleles test for this data set. Provide the p-value and the odds ratio and comment on the results.
```{r}
df <- matrix(c(112,278,150,206,348	,150),byrow=TRUE,ncol=3)
colnames(df) <- c("AA","Aa","aa")
rownames(df) <- c("Cases","Controls")
df

alleles<-cbind(2*df[,1]+df[,2],2*df[,3]+df[,2])
colnames(alleles) <- c("A","a")
fisher.test(alleles)
```
The p-value is 0.000231 and the odds ratio is 0.7406221. This means that we accept the alternative hypothesis meaning that the true odds ratio is not equal to 1. Therefore we can say that there is a trend in the proportions of the SNP and the Alzheimer's disease.

### 2. Test for association using a codominant, a dominant and a recessive model. Provide the p-values for all the tests and comment on the results.
```{r}
#Codominant
codominant <- matrix(c(112,278,150,206,348,150),byrow=TRUE,ncol=3)
colnames(codominant) <- c("AA","AB","BB")
rownames(codominant) <- c("Cases","Controls")
codominant

results<-chisq.test(codominant)
results
results$expected
fisher.test(codominant)

#Dominant
dominant <- cbind(codominant[,1],codominant[,2]+codominant[,3])
colnames(dominant) <- c("AA","AB or BB")
rownames(dominant) <- c("Cases","Control")
dominant
results<-chisq.test(dominant)
results
results<-chisq.test(codominant, correct = FALSE)
results

#Recessive
recessive <- cbind(codominant[,1]+codominant[,2],codominant[,3])
colnames(recessive) <- c("AA or AB","BB")
rownames(recessive) <- c("Cases","Control")
recessive
results<-chisq.test(recessive)
results
fisher.test(recessive)
```

### 3. Plot the risk of disease as a function of the number of m alleles. Comment on the results. Which model seems most appropriate?
```{r}
#Codominant
m_alleles <- c("AA", "AB", "BB")
risk_of_disease <- c(codominant[1,1]/(codominant[1,1]+codominant[1,2]), codominant[1,2]/(codominant[1,2]+codominant[2,2]), codominant[1,3]/(codominant[1,3]+codominant[2,3]))

data <- data.frame(m_alleles, risk_of_disease)

ggplot(data, aes(x = m_alleles, y = risk_of_disease,group = c(1,1,1))) +
  geom_point() + geom_line() +
  labs(x = "Alleles", y = "Risk of Disease") +
  ggtitle("Risk of Disease vs. Number of M Alleles") +
  theme_minimal()


#Dominant
m_alleles <- c("AA", "AB", "BB")
risk_of_disease <- c(dominant[1,1]/(dominant[1,1]+dominant[1,2]), dominant[1,2]/(dominant[1,2]+dominant[2,2]), dominant[1,2]/(dominant[1,2]+dominant[2,2]))

data <- data.frame(m_alleles, risk_of_disease)

ggplot(data, aes(x = m_alleles, y = risk_of_disease,group = c(1,1,1))) +
  geom_point() +
  geom_line() +
  labs(x = "Alleles", y = "Risk of Disease") +
  ggtitle("Risk of Disease vs. Number of M Alleles") +
  theme_minimal()

#Recessive
m_alleles <- as.factor(c("AA", "AB", "BB"))
risk_of_disease <- c(recessive[1,1]/(recessive[1,1]+recessive[1,2]), recessive[1,1]/(recessive[1,1]+recessive[2,1]), recessive[1,2]/(recessive[1,2]+recessive[2,2]))

data <- data.frame(m_alleles, risk_of_disease)

ggplot(data, aes(x = m_alleles, y = risk_of_disease,group = c(1,1,1))) +
  geom_point() +
  geom_line() + 
  labs(x = "Alleles", y = "Risk of Disease") +
  ggtitle("Risk of Disease vs. Number of M Alleles") +
  theme_minimal()
```

### 4. Perform Armitage trend test for this data set. Does the null hypothesis β1 = 0 hold? Comment on your response.
```{r}
n <- sum(df)

cas <- rep(c(0,1,2), df[1,])
con <- rep(c(0,1,2), df[2,])

y <- c(rep(1,sum(df[1,])),rep(0,sum(df[2,])))
x <- c(cas,con)

r <- cor(x,y)
A <- n*(r^2)

pval <- pchisq(A,df=1,lower.tail = FALSE)
pval
```
No, the null hypothesis doesn't hold. This means that there is a a trend in the proportions of the SNP and the Alzheimer's disease

### 5. Is there evidence for association of this marker with the disease? Argument your response
We can say there is a clear associattion between the marker and the disease. This can be clearly seen in the after the allele and Armitage test have been performed, where we got a pvalue lower than 0.05 for both tests, this means that there is a trend in the proportions of the SNP with the Alzheimers 







